Measuring causal influence with\\ back-to-back regression

### 2019-05-22 paper/meg_exps.tex
Fix
Merging...

\newpage
\subsection{Dataset}

Next, we apply our method to brain imaging data from the anonymized multimodal
neuroimaging ``Mother Of all Unification Studies'' (MOUS) dataset
(\cite{schoffelen2019204}).  The dataset contains magneto-encephalography (MEG)
recordings of 102 healthy native-Dutch adults who participated in a reading
task.
%
Subjects were exposed to a rapid serial visual presentation of Dutch words. The
word lists consisted of 120 sentences, and scrambled lists of the same words.
Each word was presented on the computer screen for 351ms on average (min: 300ms,
max: 1400ms).  Successive words were separated by a blank screen for 300ms, and
successive sentences were separated by an empty screen for a few (3-4) seconds.

\subsection{MEG preprocessing}

The raw MEG data was bandpass-filtered between 0.1 and 40 Hz using MNE-Python
default parameters (\cite{gramfort2013meg, gramfort2014mne}). Specifically, we used a zero-phase finite impulse
response filter (FIR) with a Hamming window and with transition bands of 0.1 Hz
and 10 Hz for the low and high cut-off frequencies respectively.

The raw data was then segmented 100 ms before word onset and 1.000 sec after
word onset. Time $t=0$ms corresponds to the word onset. Finally, each resulting
segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
thus led a sampling frequency of 240 Hz. We show, in Figure \ref{fig:megavg},
the average response of the magnetometers evoked by the words presented to the
first subject.

@@

\subsection{Word features} We annotated each word with 38 features: 26
correspond to the count of each letter in the word (from 'a' to 'z',
independently of their case), the total number of letters ('word length'), the
frequency of the word in natural language (using the Zipf logarithmic scale
of \cite{van2014subtlex}), and the part-of-speech tags (i.e. NOUN, PROPN (proper
noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ
(adjective), ADV (adverb), and NUM (numerical)) using the packages Spacy
(\cite{spacy2}) and WordFreq (\cite{speerwordfreq}). For clarity purposes, we refer to these
features using two broad categories: visual (i.e. each letter and the total
number of letters) and lexical (part-of-speech and word frequency).  This
procedure yields an $X \in \mathbb{R}^{n, d}$ matrix of $n\approx$ 2700 words by
$d=38$ features for each subject. Each of the columns of $X$ is normalized to
have a mean and a standard deviation of 0 and 1 respectively.

\subsection{Analyses} We attempt to identify which factors of $X$ explain the MEG
recordings $Y$ at each time sample and for each subject separately. To this aim,
we implement three types of analyses (forward, backward and back-to-back
regressions). Corresponding results can be found in Figure \ref{fig:meg_twocurves}.

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_twocurves.pdf}
  \caption{We compare the encoding approach (columns 1 and 2, ridge regression $X \rightarrow Y$) and the decoding approach (column 3, ridge regression $Y \rightarrow X$) to our method (column 4, $X \rightarrow Y \rightarrow X$) on real MEG data from the MOUS dataset. Top row: the first subject. Bottom row: average across subjects.
  We plot the contributions of the \textit{word frequency} and the \textit{word
  length} over time. Our method can disambiguate them consistently with
  previous neuroscientific work showing an early response ($\approx$ 150ms) to
  length and a later response ($\approx$ 400ms) to frequency. See text for
  details.}
  \label{fig:meg_twocurves}
\end{figure}

\paragraph{Forward regression} also known as 'encoding' in neuroimaging, here consists of
fitting a ridge-regularized ordinary least squares (OLS) regression \footnote{We use scikit-learn (\cite{sklearn}).} from $X$ to $Y$:
\begin{equation}
  H = (X^{T}X+\lambda I)^{-1} X^{T}Y
\end{equation}
where $\lambda$ is the
regularization parameter, cross-validated between $10^{-5}$ and $10^5$.

We then estimate the contribution
$diag(\hat E) \in\mathbb{D}^{d} $ of each factor. However, $H\in\mathbb{R}^{d,
q}$ is a matrix of $d$ word features by $q$ MEG channels. We approximate
the relative contribution of each feature with the sum-of-squares of the corresponding coefficients in $H$; for the $j$-th feature:
\begin{equation}
  \hat e_j = \sum_{i=0}^{q} {h_{i,j}}^2
\end{equation}
\paragraph{Backward regression} or 'decoding', is similar but predicts $X$ from $Y$:
\begin{equation} G = (Y^{T}Y+\lambda I)^{-1} Y^{T}X \end{equation}
where $\lambda$ is cross-validated as for encoding. We estimate the
contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor in the following
way. We predict $X$ on a held-out split \textit{test}:
% Using 10-fold cross-validation, we thus assessed the ability of this ridge
% regression to predict independent and identically distributed data:
\begin{equation} \hat X_{test} = Y_{test}\cdot G_{train} \end{equation}
and assess, for each factor of X independently, the Pearson correlation
coefficient $r$ between $X_{test}$ and $\hat X_{test}$. We repeat the procedure
for 20 random splits and average the result.

@@

\subsection{Second-level statistics}

The three types of analyses were implemented for each subject and each time
sample independently.  To estimate the reliability of these estimates in the
population, we tested, using one-sample t-test whether the distribution of the
estimated contribution of each factor different from baseline ($t=-100$ms):

\begin{equation} ttest(\hat E_{t=-100} - \hat E_{\tau}) \end{equation} with
$\tau \in $ 150ms and 400ms respectively

\subsection{Results}

@@

\subsection{Discussion of the results.}

>>>>

%\newpage
%\subsection{Dataset}

Next, we apply our method to brain imaging data from the anonymized multimodal
neuroimaging ``Mother Of all Unification Studies'' (MOUS) dataset
\citep{schoffelen2019204}.  The dataset contains magneto-encephalography (MEG)
recordings of 102 healthy native-Dutch adults who participated in a reading
task.
%
Subjects were exposed to a rapid serial visual presentation of Dutch words. The
word lists consisted of 120 sentences, and scrambled lists of the same words.
Each word was presented on the computer screen for 351ms on average (min: 300ms,
max: 1400ms).  Successive words were separated by a blank screen for 300ms, and
successive sentences were separated by an empty screen for a few (3-4) seconds.

\subsubsection{MEG preprocessing}

The raw MEG data was bandpass-filtered between 0.1 and 40Hz using MNE-Python
default parameters \citep{gramfort2013meg, gramfort2014mne}. Specifically, we used a zero-phase finite impulse
response filter (FIR) with a Hamming window and with transition bands of 0.1Hz
and 10Hz for the low and high cut-off frequencies respectively.

The raw data was then segmented 100ms before word onset and 1.000 sec after
word onset. Time $t=0$ms corresponds to the word onset. Finally, each resulting
segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
thus led a sampling frequency of 240Hz. Figure \ref{fig:megavg} shows
the average response of the magnetometers evoked by the words presented to the
first subject.

@@

\subsubsection{Word features} We annotated each word with 38 features: 26
correspond to the count of each letter in the word (from 'a' to 'z',
independently of their case), the total number of letters ('word length'), the
frequency of the word in natural language (using the Zipf logarithmic scale
of \citep{van2014subtlex}), and the part-of-speech tags (i.e. NOUN, PROPN (proper
noun), VERB, PRON (pronoun), CONJ (conjunction), DET (determinant), ADJ
(adjective), ADV (adverb), and NUM (numerical)) using the packages Spacy
\citep{spacy2} and WordFreq \citep{speerwordfreq}. For clarity purposes, we refer to these
features using two broad categories: visual (i.e. each letter and the total
number of letters) and lexical (part-of-speech and word frequency).  This
procedure yields an $X \in \mathbb{R}^{n, d}$ matrix of $n\approx$ 2700 words by
$d=38$ features for each subject. Each of the columns of $X$ is normalized to
have a mean and a standard deviation of 0 and 1 respectively.

\subsubsection{Methods}

We attempt to identify which factors of $X$ explain the MEG
recordings $Y$ at each time sample and for each subject separately. To this aim,
we compare three methods (forward, backward and back-to-back
regressions). Corresponding results can be found in Figure \ref{fig:meg_twocurves}.

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_twocurves.pdf}
  \caption{We compare the encoding approach (columns 1 and 2, ridge regression $X \rightarrow Y$) and the decoding approach (column 3, ridge regression $Y \rightarrow X$) to our method (column 4, $X \rightarrow Y \rightarrow X$) on real MEG data from the MOUS dataset. Top row: the first subject. Bottom row: average across subjects.
  We plot the contributions of the \textit{word frequency} and the \textit{word
  length} over time. Our method can disambiguate them consistently with
  previous neuroscientific work showing an early response ($\approx$ 150ms) to
  length and a later response ($\approx$ 400ms) to frequency.}
  \label{fig:meg_twocurves}
\end{figure}

Forward regression performs a regularized least-squares regression $A \in \mathbb{R}^{d_x \times d_y}$ from $X$ to $Y$, and estimates the causal influence $E_{i,i}$ as $\| A_{i, :} \|^2$.
%
Backward regression performs a regularized least-squares regression $B \in \mathbb{R}^{d_y \times d_x}$ from $Y$ to $X$, and estimates the causal influence $E_{i, i}$ as the Pearson correlation between the $i$-th column of $X$ and the $i$-th column of $YB$, using a validation sample.
%

% \paragraph{Forward regression} also known as 'encoding' in neuroimaging, here consists of
% fitting a ridge-regularized ordinary least squares (OLS) regression \footnote{We use scikit-learn \citep{sklearn}.} from $X$ to $Y$:
% \begin{equation}
%   H = (X^{T}X+\lambda I)^{-1} X^{T}Y
% \end{equation}
% where $\lambda$ is the
% regularization parameter, cross-validated between $10^{-5}$ and $10^5$.
% 
% We then estimate the contribution
% $diag(\hat E) \in\mathbb{D}^{d} $ of each factor. However, $H\in\mathbb{R}^{d,
% q}$ is a matrix of $d$ word features by $q$ MEG channels. We approximate
% the relative contribution of each feature with the sum-of-squares of the corresponding coefficients in $H$; for the $j$-th feature:
% \begin{equation}
%   \hat e_j = \sum_{i=0}^{q} {h_{i,j}}^2
% \end{equation}
% \paragraph{Backward regression} or 'decoding', is similar but predicts $X$ from $Y$:
% \begin{equation} G = (Y^{T}Y+\lambda I)^{-1} Y^{T}X \end{equation}
% where $\lambda$ is cross-validated as for encoding. We estimate the
% contribution $diag(\hat E) \in\mathbb{D}^{d} $ of each factor in the following
% way. We predict $X$ on a held-out split \textit{test}:
% % Using 10-fold cross-validation, we thus assessed the ability of this ridge
% % regression to predict independent and identically distributed data:
% \begin{equation} \hat X_{test} = Y_{test}\cdot G_{train} \end{equation}
% and assess, for each factor of X independently, the Pearson correlation
% coefficient $r$ between $X_{test}$ and $\hat X_{test}$. We repeat the procedure
% for 20 random splits and average the result.

@@

\subsubsection{Second-level statistics}

The three types of analyses were implemented for each subject and each time
sample independently.  To estimate the reliability of these estimates in the
population, we tested, using one-sample $t$-test whether the distribution of
the estimated contribution of each factor different from baseline ($t=-100$ms),
as $\text{t-test}(\hat E_{t=-100} - \hat E_{\tau})$, with $\tau \in $ 150ms and
400ms respectively

\subsubsection{Results}

@@

### 2019-05-22 paper/paper.tex
Merging...

If using regularized least-squares, arguably the most commonly employed linear regression technique, the solution of B2BR has a closed form solution.
%
In particular, $\hat G = (Y^\top Y + \Lambda_Y)^{-1} Y^\top X$ and $\hat H=(X^\top X + \Lambda_X)^{-1} X^\top YG$, where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices containing regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.

@@

Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
%
Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split.

@@

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{input data $X \in \mathbb{R}^{n \times d_x}$, output data $Y \in \mathbb{R}^{n\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
    \KwOut{estimate of causal influences $\hat{E} \in \mathbb{D}^{d_x \times d_x}$.}
    $\hat{E} \leftarrow 0$\;
    \For{$i = 1, \ldots, m$}{
        $(X, Y) \leftarrow \text{ShuffleRows}((X, Y))$\;
        $(X_1, Y_1), (X_2, Y_2) \leftarrow \text{SplitRowsInHalf}((X, Y))$\;
        $\hat{G} = \text{LinearRegression}(Y_1, X_1)$ \Comment*[r]{$\hat G = (Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1$}
        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$\;
        $E \leftarrow E + \text{Diag}(H)$\;
    }
    \Return{$\hat{E} / m$}
    \caption{Back-to-back regression.}
    \label{algorithm:b2br}
\end{algorithm}

@@

\subsection{Post-processing of B2BR}

@@

In regression analysis, the traditional approach to this problem consists in testing whether the regression coefficients differ from zero.
%
However, this test will not succeed here, since the estimate $\hat{E}$ is obtained from a double regression, and any employed regularization will add a bias to the diagonal of $\hat{E}$.
%
Instead, we treat the binarization of $\hat{E}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''.
%
More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
%
Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
%
To binarize $\hat{E}$, set to one all the diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the diagonal entries.

\section{Experiments}
\subsection{Synthetic data}

@@

\subsection{Related regression methods}
FC: discuss similar methods of regression, that can be used for feature extraction
And maybe have something about other causal discovery methods

>>>>

If using regularized least-squares, arguably the most commonly employed linear regression technique, the solution of B2BR has a closed form solution.
%
In particular, $\hat G = (Y^\top Y + \Lambda_Y)^{-1} Y^\top X$ and $\hat H=(X^\top X + \Lambda_X)^{-1} X^\top Y \hat G$, where $\Lambda_X$ and $\Lambda_Y$ are two diagonal matrices containing regularization parameters, useful to invert the covariance matrices of $X$ and $Y$ if these are ill-conditioned.

@@

Performing two regressions over the same data sample can result in overfitting, as spurious correlations in the data absorbed by the first regression will be leveraged by the second one.
%
To avoid this issue, we split our sample $(X, Y)$ at random into two equally-sized halves $(X_1, Y_1)$ and $(X_2, Y_2)$.
%
Then, the first regression is performed using $(X_1, Y_1)$, and the second regression is performed using $(X_2, Y_2)$.
%
To compensate for the reduction in sample size caused by the split, B2BR is repeated over many random splits, and the final estimate $\hat E$ of the causal influence matrix is the average over the estimates associated to each split.
%
After obtaining $\hat{E}$, we can fit a final regression from $X \hat{E}$ to $Y$ as a final B2BR prediction rule.

@@

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{input data $X \in \mathbb{R}^{n \times d_x}$, output data $Y \in \mathbb{R}^{n\times d_y}$, number of repetitions $m \in \mathbb{N}$.}
    \KwOut{estimate of causal influences $\hat{E} \in \mathbb{D}^{d_x \times d_x}$.}
    $\hat{E} \leftarrow 0$\;
    \For{$i = 1, \ldots, m$}{
        $(X, Y) \leftarrow \text{ShuffleRows}((X, Y))$\;
        $(X_1, Y_1), (X_2, Y_2) \leftarrow \text{SplitRowsInHalf}((X, Y))$\;
        $\hat{G} = \text{LinearRegression}(Y_1, X_1)$ \Comment*[r]{$\hat G = (Y_1^\top Y_1 + \Lambda_Y)^{-1} Y_1^\top X_1$}
        $\hat{H} = \text{LinearRegression}(X_2, Y_2 \hat{G})$ \Comment*[r]{$\hat H=(X_2^\top X_2 + \Lambda_X)^{-1} X_2^\top Y_2 \hat G$}
        $\hat{E} \leftarrow \hat{E} + \text{Diag}(H)$\;
    }
    $\hat{E} \leftarrow \hat{E} / m$\;
    $\hat{W} \leftarrow \text{LinearRegression}(X \hat{E}, Y)$\;
    \Return{$\hat{E}$, $\hat{W}$}
    \caption{Back-to-back regression.}
    \label{algorithm:b2br}
\end{algorithm}

@@

\subsection{Binarizing $\hat{E}$}

@@

In regression analysis, the traditional approach to this problem consists in testing whether the regression coefficients differ from zero.
%
However, this test will not succeed here, since the estimate $\hat{E}$ is obtained from a double regression, and any employed regularization will add a bias to the diagonal of $\hat{E}$.
%
Instead, we treat the binarization of $\hat{E}$ as a clustering problem: separate the elements in the diagonal into a group of ``small values'', and a group of ``large values''.
%
More specifically, we propose to maximize the ratio of inter-group variance and to minimize the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values.
%
Letting $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, we select the split maximizing the Sonquist-Morgan \citep{todo} criterion $\frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}$.
%
To binarize $\hat{E}$, set to one all the diagonal entries belonging to the ``large values'' group in the decided split, and setting to zero the rest of the diagonal entries.

\section{Experiments}

We perform two experiments to evaluate the performance of BB2R: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography dataset.
%
We use scikit-learn to implement all of our simulations \citep{sklearn}.

\subsection{Synthetic data}

@@
